---
title: "Data visualization and mapping"
subtitle: "Soc 225: Data & Society"
author: "[PUT YOUR NAME HERE]"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

For the first half of this lab, we'll use the `gapminder` data set. This example is adapted from Kieran Healy's book, *Data Visualization: A Practical Introduction*, chapters 3 and 4: http://socviz.co/

For the second half, we'll use the Seattle Airbnb listings from Inside Airbnb.

# Setup and data import

```{r}
library(tidyverse)
# install.packages("gapminder")
library(gapminder)

theme_set(theme_minimal())
glimpse(gapminder)
```

Read about the origins of the `gapminder` data set by typing `?gapminder`, and by looking at the source of the data: https://www.gapminder.org/data/

# Visualization: scatterplots

Scatterplots are a good way to visualize the relationship between two variables, and to look for outliers.

```{r}
p <- 
  ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point()
p
```

## Scales

To better see our data, we can transform the x-axis into a log scale.

```{r}
p + scale_x_log10()
```


## Colors

In addition to the x and y axes, another aesthetic available to use is `color`.

For example, we can color data points by continent: 

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) + 
  geom_point() + 
  scale_x_log10()
```

If you map a continuous/numeric variable onto color, `ggplot2` will pick a gradient scale: 

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = log(pop))) + 
  geom_point() + 
  scale_x_log10()
```

## Facets

Let's take a different way of breaking down these data by continent. This time, we'll *facet* the data into "small multiple" plots.

To do this, we add a new layer with `facet_wrap`. Note: the syntax is slightly different! You use a tilde (`~`) before the variable name.

(Why? Because you can facet by more than one variable. In R, this syntax is called a *formula*.)

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point() + 
  scale_x_log10() + 
  facet_wrap(~continent) 
```


## Data transformation: `mutate`

What if we wanted to look at *total* GDP, not just per capita GDP? To do this, we'd have to create a new variable, using the verb `mutate`. Total GDP is GDP per capita times population.

```{r}
gapminder %>% 
  filter(year == 2007) %>%
  mutate(gdp = gdpPercap * pop) %>%
  ggplot(aes(x = pop, y = gdp, color = continent)) + 
  geom_point()
```


## Questions: Life expectancy over time

Instead of looking at the relationship between life expectancy and GDP, now we'll look at changes in life expectancy over time. 

Create a plot where `x = year` and `y = lifeExp`. This time, use a new geom: `geom_line()` instead of `geom_point()`. 

```{r}
# write your code here!
```

a. Try creating a line plot of life expectancy by year. Initially, it won't look quite right.  

b. You need to tell the plot to group the lines by country. To do this, you'll need a new aesthetic, `group = country`. Incorporate this into your plot. Does it look more reasonable now?

c. Finally, facet by continent. Does life expectancy seem to have increased over time everywhere? Do you see any dips or decreases? 

# Maps

You can create geographical maps in R, using latitude and longitude data. This uses a new package called `leaflet` (https://rstudio.github.io/leaflet/), which you'll need to install. 

The Inside Airbnb data has latitudes and longitudes for each listing, so we'll use that.

## Points and popups

`leaflet` relies on the pipe (`%>%`) to add layers to maps. 

```{r}
# install.packages("leaflet")
library(leaflet)

airbnb_data <- read_csv("data/listings.csv")

leaflet(airbnb_data) %>%
  addTiles() %>%
  addCircles(popup = ~name)
```

## Colors and legends: qualitative/categorical

Colors are a bit less automatic than with `ggplot2`. You need to create a palette using one of the provided functions, and then use that palette for your data and the legend. 

```{r}
# this palette is based on the type of room
room_type_pal <- colorFactor("Set1", airbnb_data$room_type)

airbnb_data %>%
  filter(neighbourhood == "University District") %>%
  leaflet() %>%
  addTiles() %>%
  addCircles(popup = ~name, 
             color = ~room_type_pal(room_type)) %>%
  addLegend(pal = room_type_pal, values = ~room_type)
```

## Questions: colors for a numeric variable

You should use different color palettes for categorical vs numeric data. You've got a couple options for plotting numeric data:

- `colorNumeric`: linear mapping
- `colorBin`: bins by values
- `colorQuantile`: bins by quantiles

```{r}
# filter out outliers
# remember, you should look at those outliers first!
# and then **explain** why you're deciding to remove them
# here, we might think that some of those large values are unreliable
# they also make some of the scales harder to use and interpret
new_airbnb_data <- 
  airbnb_data %>%
  filter(price <= 500)

price_pal <- colorNumeric("RdPu", new_airbnb_data$price)

## what if you used a quantile-based approach?
## would you need to remove outliers in this case?
# price_pal <-
#   colorQuantile("RdPu", new_airbnb_data$price, n = 5)

## uncomment this code, and use it as a scaffold 
## for coloring the data points by price
# leaflet(new_airbnb_data) %>%
#   addTiles() %>%
#   addCircles() %>%
#   addLegend()
```

a. How many outliers did you filter out? Does that seem reasonable to you, or misleading? 

b. Un-comment the skeleton of code above, and modify it to color the data points by price. Try both the value-based and quantile-based approaches. 

c. Look at the map and explore it. Name one area of the city that seems to have higher-than-average prices, and one area that seems to have lower-than-average prices.

d. Which mapping of prices to colors---linear mapping by value or binning by quantile---do you think is more useful here, and why? There isn't a right answer.

# Challenge: Apply what you've learned

You've learned a variety of techniques for data transformation and visualization: 

- filtering and mutating
- `geom_point`, `geom_line`, `geom_bar`, and `geom_histogram`
- colors, facets, and groups

Apply these tools to the Airbnb data set to visualize price and/or review information. Break the data down in some informative way, e.g. by neighbourhood group or room type. Try a few different iterations, and interpret what you see. 
