---
title: "Data visualization and mapping"
subtitle: "Soc 225: Data & Society"
author: "[PUT YOUR NAME HERE]"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

# Goals

- Add color to our visualizations
- Add facets to our visualizations
- Make visualizations using maps
- Control color output using palettes

# 0: Lab Check-in

For the first half of this lab, we'll use the `gapminder` data set. This example is adapted from Kieran Healy's book, *Data Visualization: A Practical Introduction*, chapters 3 and 4: http://socviz.co/

For the second half, we'll use the Seattle Airbnb listings from Inside Airbnb.

** Question 0.1: Use the console to install the package 'gapminder' package, load that package and the tidyverse package.**

```{r}

```

# 1 Visualizations with Covariation

Bar charts and histograms are great at showing variation along one variable. However, we're often interesting at the relationship between two or more variables. That is, a bar chart or histogram shows the *variation* of a sigle variable. The visualizations we'll learn in the module show the *covariation* between variables.

Before we start, let's take a look at the gapminder data

** Question 1.1: Use glimpse to examine the gapminder data **

```{r}

```

** Question 1.2: What's one variable that is continuous? What's one that's categorical? **

Read about the origins of the `gapminder` data set by typing `?gapminder`, and by looking at the source of the data: https://www.gapminder.org/data/

## Scatterplots

Scatterplots are a good way to visualize the relationship between two variables, and to look for outliers.

```{r}
theme_set(theme_minimal()) # remember that I like the minimal theme? I can set it as default

p <- # I can use the assignment operator to save a plot!
  ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) + # use the gapminder data, set x to the gdp and y to the life expectancy
  geom_point() # plot points (x,y) for each country in the dataset

p # by calling the object 'p' we can display the graph
```

** Question 1.3 Make a scatterplot comparing the population and the life expectancy, assign it to an object named 'q'**

```{r}

```

## Scales

To better see our data, we can transform the x-axis into a log scale.

```{r}
p + # taking our old graph, the '+' lets us add to it
  scale_x_log10() # add our scale
```

** Question 1.4 add a log scale to your graph q **

```{r}

```

** Question 1.5 what do you think accounts for the distinctive shape of your scatterplot? **

## Colors

In addition to the x and y axes, another aesthetic available to use is `color`. This means we can look at how up to three variables change together!

For example, we can color data points by continent, a categorical variable: 

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) + # same as p above but adding color
  geom_point() + # plot the points
  scale_x_log10() # add the same scale
```

If you map a continuous/numeric variable onto color, `ggplot2` will pick a gradient scale: 

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = log(pop))) + 
  geom_point() + 
  scale_x_log10()
```

** Question 1.6 Remake plot q by choosing a continuous variable to add colors **

```{r}

```

** Question 1.7 Remake plot q by choosing a categorical variable to add colors that explain the interesting shapes **

```{r}

```


## Facets

Let's take a different way of breaking down these data by continent. This time, we'll *facet* the data into "small multiple" plots.

To do this, we add a new layer with `facet_wrap`. Note: the syntax is slightly different! You use a tilde (`~`) before the variable name.

(Why? Because you can facet by more than one variable. In R, this syntax is called a *formula*.)

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) + 
  geom_point() + 
  scale_x_log10() + 
  facet_wrap(~continent) 
```

** Question 1.8: Add a continent facet wrap to q **

```{r}

```


# 2: Make new columns: `mutate`

The `mutate` verb is one of the most important. It allows you to add columns (variables) to your dataset. As its name suggests, the most common way of doing that is by adjusting, or mutating, existing variables.

What if we wanted to look at *total* GDP, not just per capita GDP? To do this, we'd have to create a new variable, using the verb `mutate`. Total GDP is GDP per capita times population.

```{r}
gapminder %>% # take the gapminder data
  filter(year == 2007) %>% # filter to the year 2007
  mutate(gdp = gdpPercap * pop) %>% # make a new column 'gdp' which is the product of percapita gdp and population
  ggplot(aes(x = pop, y = gdp, color = continent)) + # start a plot with the x axis as population and gdp as the y, color by contient
  geom_point() # plot the points
```


# 3 Exercise: Life expectancy over time

Instead of looking at the relationship between life expectancy and GDP, now we'll look at changes in life expectancy over time. You can use the last code block for all the questions.

## 3.1: Line plot of life expectancy by year 

**Create a plot where `x = year` and `y = lifeExp`.** This time, use a new geom: `geom_line()` instead of `geom_point()`. Initially, it won't look quite right.  


## 3.2: The `group` aesthetic 

**You need to tell the plot to group the lines by country.** To do this, you'll need a new aesthetic, `group = country`. Incorporate this into your plot. Does it look more reasonable now?

## 3.3: Facet and interpret

**Finally, facet by continent.** Does life expectancy seem to have increased over time everywhere? Do you see any dips or decreases? 

```{r}

```


# 4: Maps

You can create maps in R using latitude and longitude data. This uses a new package called `leaflet` (https://rstudio.github.io/leaflet/), which you'll need to install in the console. 

The Inside Airbnb data has latitudes and longitudes for each listing, so we'll use that.

**Question 4.1: Install leaflet, load it, and then read the file 'data/listings.csv' and name it airbnb_data**

```{r}

```


## Points and popups

`leaflet` relies on the pipe (`%>%`) to add layers to maps. 

```{r}
leaflet(airbnb_data) %>% # begin by passing the data to leaflet
  addTiles() %>% # add the map files to the plot, leaflet automattically uses the latitdue and longitude data to find the right map
  addCircles(popup = ~name) # add circles for each listing, by adding the popup argument, we can click on a cirle to show the name
```

**Question 4.2: Make a leaflet plot that only includes only listings with a price over $200 and shows the price when the circle is clicked**

```{r}

```

## Colors and legends: qualitative/categorical

Colors are a bit less automatic than with `ggplot2`. You need to create a palette using one of the provided functions, and then use that palette for your data and the legend. 

```{r}
# We're going to start by making a smaller dataframe to use for our visualization
example_data <- 
  airbnb_data %>% # assign the small dataframe the name 'example_data'
  filter(neighbourhood == "University District") # filter the example data to only include listings from the U district
 
# this palette is based on the type of room
room_type_pal <- colorFactor("Set1", example_data$room_type) # make a palette with the room types from the example data

leaflet(example_data) %>% # make a leaflet plot with our example data
  addTiles() %>% # add the map tiles
  addCircles(popup = ~name, # add circles which we can click to see the names
             color = ~room_type_pal(room_type)) %>% # color the circles with our palette
  addLegend(pal = room_type_pal, values = ~room_type) # add a legend so we know what's what
```



**Question 4.3: Use the same code from above, but map a neighborhood group of your choice**

```{r}

```


## Colors and legends: numeric/continuous

You should use different color palettes for categorical vs numeric data. You've got a couple options for plotting numeric data:

- `colorNumeric`: linear mapping of numbers onto a color gradient. For example, `colorNumeric("RdPu", data$variable)`.

- `colorBin`: bins by values, so each color spans the same numeric range. For example, `colorBin("RdPu", data$variable, bins = 5)`.  

- `colorQuantile`: bins by quantiles, so each color has the same number of data points. For example, `colorQuantile("RdPu", data$variable, n = 5)`.

The examples above use a red-purple gradient ("RdPu") as the color palette. Type `RColorBrewer::brewer.pal.info` into your console for a full list of possible palettes.

```{r}
example_data <- 
  airbnb_data %>% 
  filter(neighbourhood == "University District") 

price_pal <- colorNumeric("RdPu", example_data$price)

leaflet(example_data) %>%
  addTiles() %>%
  addCircles(popup = ~name, 
             color = ~price_pal(price)) %>%
  addLegend(pal = price_pal, values = ~price)
```

**Question 4.4: Use the same code from above, but make a palette based on quantiles instead.**

```{r}

```

**Question 4.5: Compare the two maps of prices in the UDistrict. Which mapping of prices to colors---linear mapping by value or binning by quantile---do you think is more useful here, and why? There isn't a right answer.**





# 5 Exercise: Mapping Airbnb in Seattle

## 5.1: Choose a variable

**Think about what variable you'd like to display in your map.**

[Write your variable name here]


## 5.2: Plot the distribution 

**Create a bar plot or histogram as appropriate to look at the distribution of your variable**

```{r}

```

## 5.3 Are there values that seem like outliers? 

Outliers are data points that seem very different from the rest of the data. For instance, if one listing costs 10x the average price, then a color palette using linear mapping will show that listing as one color and all the rest of the listings as another. In other words, outliers can make it hard to see important differences (they can cause other problems for analysis too). That might be a good reason to exclude those listings from our map. On the other hand, removing outliers might remove important data: don't we want the people looking at our visualization to know that there was a very expensive listing? Using a quantile palette is a way to include outliers but still show variation. 

A couple reasons to think about data in this way: 

- it's part of understanding **patterns and distributions** in your data
- it helps you think about **data quality and potential flaws** in your data
- it makes you articulate your **choices and goals** when you present your data 

## 5.4: Including outliers 

**If there are outliers, do you think you should include them in your graph? Justify your answer.**



All of the subsequent code can be written in the code chunk below.

## 5.5: Create a data frame

**Create a new data frame for the rest of the exercise, filtering as necessary**

## 5.6: Create a color palette

**Use that data frame to create a color palette for your variable**

## 5.7: Create a leaflet map 

**Create a leaflet plot using your new data set**

```{r}

```


# Hints:

1.2 <int> and <dbl> are number columns, <fct> are 'factor' columns

4.1 remember that you can use unique(some_data$some_column) to list all of the unique values of some column in some data frame
