---
title: "Reading data"
subtitle: "Soc 225: Data & Society"
author: "[PUT YOUR NAME HERE]"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

# Opening this file

You should click on `01-reading-data.Rproj` when you open this lab project. Check your *working directory* if you run into trouble loading your data.

# Rectangular data

First, we'll look at rectangular data, which are data organized into a table of rows and columns. This is the most common type of data we'll see, and if data are in another form we'll often want to make them rectangular.

Check out the data folder. There's a comma-separated values (CSV) file called `seattle_airbnb.csv`. This contains data about 100 Airbnb listings from Seattle. 

These data come from Inside Airbnb, http://insideairbnb.com/. Go to the website and have a look at the About, Behind, and Get the Data pages. 

**Question: Who created these data sets? Why did they do it? Where were the data sourced from?** 



[WRITE AN ANSWER HERE]




## Setup

R doesn't come with everything we need loaded by default. Before we do anything else, we need to load a *package*. The package we'll use is called "tidyverse." You can read more about the tidyverse here: https://www.tidyverse.org/

If a package hasn't been installed on your machine, you'll need to install it. For instance, you'd type `install.packages("tidyverse")`. 

Run this chunk of code to load the tidyverse:

```{r}
library(tidyverse)
```

## Importing data

To use data inside R, we first have to import, or *read*, that data into our environment.

```{r}
airbnb_data <- read_csv("data/seattle_airbnb.csv")
```

When we do this, we assign the values of the data to a variable, `airbnb_data`, using the arrow (`<-`). The object we create is called a data frame.

## Looking at the data

You can print the data frame to the console by typing the name of the object:

```{r}
airbnb_data
```

You can also look at the entire data set using RStudio's built-in viewer: 

```{r}
# "un-comment" the line below this one, by removing the '#' and the space
# View(airbnb_data)
```

You can also look at parts of the data frame instead: 

```{r}
head(airbnb_data)
```

**Question: `head` shows the first 6 rows by default. Change the following code to show the first 10 rows:** 

```{r}
head(airbnb_data, n = 6)
```

**Question: What if you want to look at the *last* several rows of a data frame instead of the first several rows?**

To find out, read the documentation for `head` by typing `?head` into the console.

```{r}
# [TYPE YOUR ANSWER HERE, AND RUN IT]
```

You can extract a single column by name using `$`:

```{r}
head(airbnb_data$name)
```

# Hierarchical data

Data isn't always a single, flat table. Sometimes it's nested or hierarchical.

`colors.json` is a file in JSON format. This is a common format for web data. Open it up and have a look at it. You can read it into R using another package: 

```{r}
library(jsonlite)
json_data <- read_json("data/colors.json")
```

When you do, it becomes a different type of R object, a *list*: 

```{r}
str(json_data)
```

It's a little trickier, but you can use `[[]]` and `$` to extract pieces of this data. For example, the 5th color: 

```{r}
json_data$colors[[5]]
```

**Question: see if you can extract the hex code for red. Look at the output above for hints.**

```{r}
# [TYPE YOUR ANSWER HERE, AND RUN IT]
```

# Challenge: Excel worksheets

In the data folder, there is an Excel spreadsheet, `airbnb.xlsx`. It contains data for three cities (Seattle, Boston, Chicago) in separate sheets. Use the internet to find a package that will allow you to read in all of this data. If you read in each sheet separately, combine them into one data frame.

# Just for fun: Image data

Images are data too, and R can import them as well. If you have extra time, check out the vignette for the `magick` package: 

https://cran.r-project.org/web/packages/magick/vignettes/intro.html

```{r}
# install.packages("magick")
# library(magick)
# cat <- image_read("data/Black_white_cat_on_fence.jpg")
# cat
# image_flip(cat)
```

# Turning in your assignment

Press the "knit" button to produce an html file that will have all of your output. Please turn in **both** the Rmd and html files on Canvas.