---
title: "Categorical and text data"
subtitle: "Soc 225: Data & Society"
author: "[PUT YOUR NAME HERE]"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

# Goals

- Understand different data types
- Move between categorical and other types of data
- Locate keywords in texts
- Use regular expressions to manipulate text data


# 1: Modes and Types of Data in R  

We've already done a lot with lots of data here in R. But it's time to pause briefly to talk about the different kinds of data R can hold and the different formats R can hold the data in.  

## Modes

There are three kinds of data we deal with in R. We call them *modes* of data:  

Numeric: 1000, 2.3, or 4.4e23  
Character: "super cool", "airbnb is totally hard to deal with","1,000,000", or "free cheese in the break room"  
Logical: TRUE or FALSE  

**Question 1.1: assign the following three variables values of the proper mode**  

```{r}
# a <- 
# b <- 
# c <- 
```

We can check the mode of a variable by using the function `typeof()`.   

```{r}
typeof(1000)
typeof("super cool")
typeof(TRUE)
```

** Question 1.2 Check the modes of 'a', 'b', and 'c' from 1.1 using `typeof()`**  

```{r}

```

## Data Structures  

So far in this module, each of our variables has contained only one value. But, since you've worked with data frames, you know that there are many ways to organize data in R. We'll talk about six data structures.  

1. Scalar: single values  
- assign a single value to make a scalar
```{r}
my_scalar <- 45
```

2. Vector: one dimensional data type, only one mode of data allowed  
- use `c()` to make a vector:
```{r}
my_vector <- c(45, 32, 31, 10000)
```

** Question 1.3 Check the length of my_vector and my_scalar using length(). What does that tell you about scalars? **
```{r}

```

Vectors can only accept one mode of data, so if we mix modes, the vector will coerce (force) all of them to be the same:
```{r}
my_new_vector <- c(45, 4.5, "450")
typeof(my_new_vector)
```
In this case, even though 45 and 4.5 are numeric scalars, they've been coerced to character mode.

3. Matrix: two dimensional data type, only one mode of data allowed  

```{r}
my_matrix <- matrix(1:20, nrow=5,ncol=4)
my_matrix
typeof(my_matrix)
```

** Question 1.4 What data mode is filled in this matrix? Confirm your answer using typeof() **

4. Array: multi-dimensional data type, only one mode of data allowed  

If you need more than two dimensions, you'll need an array. We won't cover them, but you can look up the documentation.  

--- break to talk about getting stuff out of these data-types ---

For vectors, matricies, and arrays, accessing the data you store in them is as simple as []. That is, you just need to throw brackets around the location of the value you want and you'll get it. Each value in one of these data types has a kind of 'address' made up of a numeric vector with lenght equal to the number of dimensions. 

```{r}
# grab the third item in my_new_vector
my_new_vector[3]

# grab the item in the third row and second colum of my_matrix
my_matrix[3,2]

# grab the whole second row of my_matrix
my_matrix[,2]
```


5. List: one dimensional data type, ordered and named, *multiple data modes and types allowed*  

Lists are actually a special type of vector, but they can hold basically anything in them. You can have lists of numbers, lists, lists of matricies, lists of vegetables, whatever.

```{r}
my_list <- list(45, 4.5, "450", c(1,2,3))
my_list
```

Notice that the values were not coerced into the same mode, but were allowed to stay the same. 

Accessing things you've stored in lists is really similar to vectors. You can us [] around the index of whatevery you want to grab. However, you'll automatically be given a list of that item. To get just the item us [[]]:

```{r}
my_list[1]
my_list[[1]]
```

You can do other cool suff with lists, like add names to each item to make a named list. You can find out more by reading the help entries for list() and names().

** Question 1.5: Make a list in R of your favorite things. **

6. Data Frames: multi-dimensional (usually two dimensional) data type, only one mode of data per column, multiple per row  

This is what we will, and what we have already, used most often. Data frames resemble matricies and arrays superficially -- all three are data organized by dimension. However, Data Frames are a bit different. The columns are vectors of a single mode, but the mode of each column can be different. That means that each row is made up of a single entry from each column, and so can include data of different modes. That's great for us because we usually want to work with data where each observation (row) has variables of different types.

When we use read_csv to read a file, the package `readr` automatically makes a data frame for us. But we can also make a new data frame out of vectors. Imagine if I had some data on trees:

```{r}
tree_id <- c(1,2,3,4) # set tree ids from 1-4
tree_heights <- c(19,20,35,5) # real measured tree hights that I made with a measuring tape
tree_ages <- c(40, 35, 60, 4) # I asked each tree
tree_names <- c('oak', 'poplar', 'willow', 'beech') # two of these trees really like Pokemon
data_frame(tree_id,tree_heights,tree_ages,tree_names)
```

Then we can use all the verbs we've used from the tidyverse. 

** Question 1.6 Use mutate() to add a column that shows the average number of feet each tree has grown for each year it's been alive **

```{r}

```


# 2: Categorical data

There's actually another important mode of data in R. Sometimes we have a long vector, maybe thousands or millions of entries long, but it only has a few values. For instance, the following code makes a character vector of 1024 tree names, but it only has 4 types of trees.

```{r}
lots_of_trees <- sample(tree_names, 1024, replace=TRUE) # sample repeatedly from our list of tree names 1024 times
head(lots_of_trees) # show just the first 6 entries
```

R has a way of dealing with this kind of categorical data, called a *factor*. Factors make some kinds of statistical analysis and visualizations much easier. We're going to use a new package, called forcats, (for categorical variables, not for kitties), so install and load it.

```{r}

```

Forcats lets you turn a vector of any mode into a factor with the functions 'as_factor()'.

** Question 2.1: Turn lots_of_trees into a factor called tree_factor **
```{r}
tree_factor <- as_factor(lots_of_trees)
```

A factor has levels:
```{r}
levels(tree_factor)
```



# Text data

```{r}
library(stringr)
```
