---
title: "Categorical and text data"
subtitle: "Soc 225: Data & Society"
author: "[PUT YOUR NAME HERE]"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

# Goals

- Understand different data types
- Move between categorical and other types of data
- Locate keywords in texts
- Use regular expressions to manipulate text data

# 0: Lab Check-in

# 1: Data in R  

We've already done a lot with lots of data here in R. But it's time to pause briefly to talk about the different kinds of data R can hold and the different formats R can hold the data in.  

## Data Types

There are three main kinds of data we deal with in R. We call them *types* of data:  

Numeric (double or integer): 1000, 2.3, or 4.4e23  
Character: "super cool", "airbnb is totally hard to deal with", "1,000,000", or "free cheese in the break room"  
Logical: TRUE or FALSE  

**Question 1.1: assign the following three variables values of the proper type**  

```{r}
# a <- 
# b <- 
# c <- 
```

We can check the type of a variable by using the function `typeof()`.   

```{r}
typeof(1000)
typeof("super cool")
typeof(TRUE)
```

**Question 1.2 Check the types of 'a', 'b', and 'c' from 1.1 using `typeof()`**  

```{r}

```

## Data Structures  

So far in this module, each of our variables has contained only one value. But, since you've worked with data frames, you know that there are many ways to organize data in R. We'll talk about six data structures.  

1. Scalar: single values  
- assign a single value to make a scalar

```{r}
my_scalar <- 45
```

2. Vector: one dimensional data structure, only one type of data allowed  
- use `c()` to make a vector:

```{r}
my_vector <- c(45, 32, 31, 10000)
```

**Question 1.3 Check the length of my_vector and my_scalar using length(). What does that tell you about scalars?**

```{r}

```

Vectors can only accept one type of data, so if we mix types, the vector will coerce (force) all of them to be the same:

```{r}
my_new_vector <- c(45, 4.5, "450")
typeof(my_new_vector)
```

In this case, even though 45 and 4.5 are numeric scalars, they've been coerced to character type.

3. Matrix: two dimensional data structure, only one type of data allowed  

```{r}
my_matrix <- matrix(1:20, nrow=5,ncol=4)
my_matrix
typeof(my_matrix)
```

**Question 1.4 What data type is filled in this matrix? Confirm your answer using typeof()**

4. Array: multi-dimensional data structure, only one type of data allowed  

If you need more than two dimensions, you'll need an array. We won't cover them, but you can look up the documentation.  

--- break to talk about getting stuff out of these data-types ---

For vectors, matrices, and arrays, accessing the data you store in them is as simple as []. That is, you just need to throw brackets around the location of the value you want and you'll get it. Each value in one of these data types has a kind of 'address' made up of a numeric vector with lenght equal to the number of dimensions. 

```{r}
# grab the third item in my_new_vector
my_new_vector[3]

# grab the item in the third row and second colum of my_matrix
my_matrix[3, 2]

# grab the whole second row of my_matrix
my_matrix[, 2]
```


5. List: one dimensional data structure, ordered and named, *multiple data types allowed*  

Lists are actually a special type of vector, but they can hold basically anything in them. You can have lists of numbers, lists, lists of matrices, lists of vegetables, whatever.

```{r}
my_list <- list(45, 4.5, "450", c(1, 2, 3))
my_list
```

Notice that the values were not coerced into the same type, but were allowed to stay the same. 

Accessing things you've stored in lists is really similar to vectors. You can use [] around the index of whatever you want to grab. However, you'll automatically be given a list of that item. To get just the item use [[]]:

```{r}
my_list[1]
my_list[[1]]
```

You can do other cool stuff with lists, like add names to each item to make a named list. You can find out more by reading the help entries for list() and names().

**Question 1.5: Make a list in R of your favorite things.**

6. Data frames: two-dimensional data structure, only one type of data per column, multiple per row  

This is the data structure we use most often for data analysis in R. Data frames resemble matrices and arrays superficially -- all three are data organized into multiple dimensions. However, data frames are a bit different. The columns are vectors of a single type, but the type of each column can be different. That means that each row is made up of a single entry from each column, and so can include data of different type. That's great for us because we usually want to work with data where each observation (row) has variables of different types.

When we use `read_csv` to read a file, the package `readr` automatically makes a data frame for us. But we can also make a new data frame out of vectors. Imagine if I had some data on trees:

```{r}
library(tidyverse)

tree_id <- c(1, 2, 3, 4) # set tree ids from 1-4
tree_heights <- c(19, 20, 35, 5) # real measured tree heights that I made with a measuring tape
tree_ages <- c(40, 35, 60, 4) # I asked each tree
tree_names <- c("oak", "poplar", "willow", "beech") # two of these trees really like Pokemon

tree_data <- data_frame(tree_id, tree_heights, tree_ages, tree_names)
tree_data
```

Then we can use all the verbs we've used from the tidyverse. 

**Question 1.6: Use mutate() to add a column that shows the average number of feet each tree has grown for each year it's been alive**

```{r}

```


# 2: Categorical data (aka "factors")

Let's think back to the kinds of variables we encounter during analysis, and how we might use R's data types and data structures to represent them. Continuous variables, of course, can be represented with numeric vectors. 

But what about categorical variables? We might represent them with character vectors. But characters are open-ended and potentially messy, as you'll see when we talk about text data below.   

To better represent categorical data, R can instead use an augmented data type called a *factor*. Factors build on the basic data types you learned about above. They give us more tools specifically for working with categorical data, which are especially handy for visualizations. 

To work with factors, we're going to use a new package called `forcats`, (for categorical variables, not for kitties). `forcats` is installed as part of the `tidyverse`, but you'll need to load it separately. 

This module is based on a section of UBC's STAT 545 course (http://stat545.com/block029_factors.html) and Chapter 15 of R for Data Science (http://r4ds.had.co.nz/factors.html). 

## Example: Gapminder

We'll use the `gapminder` data set again to illustrate working with factors.

**Question 2.1: Load the forcats and gapminder packages. Look at the gapminder data. Which variables are factors?**

```{r}

```

The values that a factor can take on are called *levels*. You can use the `levels()` function to look at the levels of a factor.

**Question 2.2: Look at the levels for one of the variables in the gapminder data. How many levels does it have? What order are they in?**

```{r}

```

### Ordering factors by frequency

The gapminder data includes countries with complete data from 1952 to 2007. How many countries is that per continent? We'll start with a simple bar chart, and then go through ways to improve it step by step. 

```{r}
gapminder %>%
  # filter to one year, it doesn't matter which
  filter(year == "2007") %>%
  ggplot(aes(x = continent)) +
  geom_bar() 
```

First, we'll order the bars by the number of countries, using `fct_infreq()`: 

```{r}
gapminder %>%
  filter(year == "2007") %>%
  mutate(continent = fct_infreq(continent)) %>%
  ggplot(aes(x = continent)) + 
  geom_bar()

# more concise, but messier: 
# gapminder %>%
#   filter(year == "2007") %>%
#   ggplot(aes(x = fct_infreq(continent))) + 
#   geom_bar()
```

Next, we'll put the bars in ascending order. To reverse the order of a factor, use `fct_rev()`:

```{r}
gapminder %>%
  filter(year == "2007") %>%
  mutate(continent = fct_rev(fct_infreq(continent))) %>%
  ggplot(aes(x = continent)) + 
  geom_bar()
```

Finally, let's polish our plot a little. We'll flip the axes and add labels: 

```{r}
gapminder %>%
  filter(year == "2007") %>%
  mutate(continent = fct_rev(fct_infreq(continent))) %>%
  ggplot(aes(x = continent)) + 
  geom_bar() +
  coord_flip() + 
  labs(title = "How many countries per continent?", 
       subtitle = "Countries with complete data, Gapminder")
```

Compare this to the bar chart we started with---it should be easier now to see the point of the plot and to compare across continents.

### Ordering factors by another variable

Instead of frequencies, you can reorder factors by values of some other numeric variable using `fct_reorder()`.

The syntax is `fct_reorder(categorical_variable, continuous_variable)`. To practice this, we'll use gapminder data for Asian countries in 2007.

**Question 2.3: Create a new data frame by filtering the `gapminder` data down to countries in Asia in 2007.**

```{r}
# gapminder_asia_2007 <- 
```

**Question 2.4: Choose one of the continuous variables. Make a dot plot of this variable by country.**

You'll want country names on the y axis for legibility. With `geom_point()`, you don't need to use `coord_flip()` to do this.

```{r}

```

**Question 2.5: Reorder countries by that same continuous variable, and remake your plot.**

Be sure to give your plot an appropriate title.

```{r}

```

## Converting from characters to factors

If you want to use the tools of `forcats` on categorical data stored as a character rather than a factor, you'll need to *convert* that variable to a factor first. `forcats` will often do this for you, but you can turn a character variable into a factor explicitly using the function `as_factor()`.

**Question 2.6: Try this out on the `tree_data` from above. Convert the variable `tree_names` into a factor.**

```{r}

```

**Question 2.7: Plot one of the continuous variables against tree_names, reordering appropriately.**

Try out using `geom_col()` for this plot. How is it different from `geom_bar()`?

```{r}

```

## Going further with `forcats`

`forcats` includes many other tools for working with factors. For example, you can rename or combine factor levels using `fct_recode`, or group together small categories using `fct_lump`. Type `fct_` into your console and hit "tab" to check out the full list of forcats functions.

If you have extra time, have a look the `gss_cat` data set, included in the `forcats` package. This is some sample data from the General Social Survey (GSS), one of the most important ongoing social scientific surveys. The sample includes data on race, religion, political party affiliation, and marital status. Experiment with these other `forcats` functions on `gss_cat`.

# 3: Text data (aka "strings")

Character vectors are more than just categories---they can contain rich, messy, unstructured data that we might use to produce categories or quantities. 

When we talk about manipulating character data in this way, we usually call the data a *string*. The `stringr` package contains tools for working with strings. Like `forcats`, `stringr` is a part of the tidyverse that isn't loaded by default. It contains some sample data we'll use, a list of names of fruit.

**Question 3.1: Load the `stringr` package and check out the list of `fruit` it contains**

```{r}

```

## Combining strings

`str_c` is our basic way of joining strings together: 

```{r}
str_c("pine", "apple")
```

You can join as many as you'd like: 

```{r}
str_c("pine", "apple", "s")
```

You can see that they're joined together with no space in between. If you want to change that, use the argument `sep`:

```{r}
str_c("crab", "apple", sep = " ")
```

`str_c` is *vectorized*; it will recycle strings, like so: 

```{r}
str_c(c("pine", "crab"), "apple")
```

The argument `collapse` will combine all of the input vectors at the very end: 

```{r}
str_c(c("pine", "crab"), "apple", collapse = " ")
```

**Question 3.2: Combine all of the fruit in `fruit` into a single string, delineated by commas. Do you need to use `sep` or `collapse`? Why?**

```{r}

```

## Pulling out parts of strings

The basic idea here is that we take a **string** (or vector of strings) and apply some sort of **pattern** to it. There are different kinds of things we might want to get back out of this, depending on our goal. 

For instance, we might want 

- only the items that match the pattern (`str_subset`)
- TRUE or FALSE for every item (`str_detect`)
- the pattern itself for every item (`str_extract`)

This can be a handy way filter a data set or to create new variables out of messy text data.

Let's begin by looking at fruits starting with the letter "a". The "^" here means the beginning of a string. 

```{r}
str_subset(fruit, "^a")
```

What do we get back if we try `str_detect` with the same pattern?

```{r}
str_detect(fruit, "^a")
```

How could we use this? If we have a data frame of fruit, then `str_detect` works nicely with `filter`: 

```{r}
fruit_data <- data_frame(fruit)

fruit_data %>%
  filter(str_detect(fruit, "^a"))
```

**Question 3.3: "$" indicates the *end* of a string. Find all the fruit names that end with "a". Do this with both `fruit` and `fruit_data`.**

```{r}

```

Let's pretend I'm a fruit marketer interested in the latest trends in fruit names. I can extract particular patterns of interest using `str_extract`. (Remember, "|" means "or".)

```{r}
str_extract(fruit, "fruit|berry|melon")
```

**Question 3.4: Make a bar chart using these extracted elements. You can add any other elements that you're interested in to the pattern.**

```{r}

```

## Exercises

Type `str_` into your console to see more of the tools `stringr` contains. These exercises are meant to familiarize you with a few of them. Other functions you might find particularly useful include `str_replace` (find a pattern and replace it with something else) and `str_to_lower` (turn everything to lower case). 

**Question 3.5: Substrings. `str_sub` extracts a part of a string based on position. Check out the help for it, and use it to extract the first letter of each `fruit`. Make a bar chart of initial letter frequencies.** 

(Remember, you can use categorical tools like fct_rev and fct_infreq to improve your chart.)

```{r}

```

**Question 3.6: Splitting strings. Try out `str_split` and `str_split_fixed` on the `best_coast` data below. What's the difference in the resulting data structures?** 

For a challenge, see if you can turn the city and state into separate columns of a data frame. 

```{r}
best_coast <- c("Seattle, WA", "Portland, OR", "San Francisco, CA", "Vancouver, BC") 


```

## Going further with text data

The text module is based off of http://stat545.com/block028_character-data.html and Chapter 14 of R for Data Science (http://r4ds.had.co.nz/strings.html). You can start with those resources for a deeper dive into strings.

If you want to get fancy with extracting bits of strings, you'll need to learn more about **regex**. Regex, *regular expressions*, is a way of representing more complex patterns.  

It's possible to go much further with analyzing unstructured text data. For instance, what if you want to analyze word counts or sentiments in a novel, or a bunch of tweets? The `tidytext` package is one framework for doing this. *Text Mining with R*, by Julia Silge and David Robinson, is a free online textbook for working with tidy text data, which you can find here: https://www.tidytextmining.com/

# Hints

3.4 Start with `fruit_data` and use `mutate` to create a new variable.

3.6 Here's one way to do it. We haven't talked about the `purrr` package in the Tidyverse, but it's got some powerful tools like `map_chr`.

```
data_frame(best_coast) %>%
  mutate(split = str_split(best_coast, ", "), 
         city = map_chr(split, 1), 
         state = map_chr(split, 2)) %>%
  select(-split)
```
