---
title: "Social Networks"
subtitle: "Soc 225: Data & Society"
author: "[PUT YOUR NAME HERE]"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

# Social networks

These exercises are based off of a workshop written and taught by Alex Hanna, Pablo Barberá, and Dan Cervone: https://github.com/pablobarbera/data-science-workshop/tree/master/sna

# 0: Setup

We'll use the `tidygraph` and `ggraph` packages to manipulate and visualize networks. `tidygraph` is a tidy interface built on top of the `igraph` package, and `ggraph` is an extension of `ggplot2` for plotting network data.

You can read more about `tidygraph` here: https://www.data-imaginist.com/2017/introducing-tidygraph/

You'll need to install and load these packages: 

```{r warning=FALSE, message=FALSE}
# install.packages("tidygraph")
# install.packages("ggraph")

library(tidyverse)
library(tidygraph)
library(igraph)
# if you get an error below, reinstall ggplot2
# install.packages("ggplot2")
library(ggraph)
```

# 1: Star Wars

Our first example is a small network of character interactions in Star Wars: Episode IV, described further here: https://cdn.rawgit.com/pablobarbera/data-science-workshop/master/sna/01_networks_intro.html

## Look at the data: nodes and edges

First, read in the data. There are *two* data sets, one of nodes, and one of edges.

```{r}
nodes_sw <- read_csv("data/star-wars-network-nodes.csv") 
edges_sw <- read_csv("data/star-wars-network-edges.csv") 
```

**Look at both data sets and answer these questions:**

1.1: How many characters are there? Which data set lists the characters?

1.2: How many connections between the characters? Which data set lists the connections?

1.3: Based on that, in your own words, what do you think a *node* is? What's an *edge*?

## Make a graph

When we combine the nodes and edges, we get a network object called a *graph*. 

- `igraph::graph_from_data_frame` does the heavy lifting, turning one or more data frames into a graph
- `tidygraph::as_tbl_graph` makes that graph easier to work with and plot

We connect them, as usual, with the pipe (`%>%`).

```{r}
g_sw <- 
  graph_from_data_frame(d = edges_sw, vertices = nodes_sw, directed = FALSE) %>%
  as_tbl_graph()
```

Type `g_sw` into your console and see what prints out.

**1.4: This is an *undirected* graph. Why?**

## Plotting a graph

To make sense of network data, it's good to visualize it. `ggraph` is like `ggplot` but for tidy graph objects.

What's different?

- You have to think about *layouts*. Where should the different nodes go? 

  There are "force-directed" algorithms to help with this. Because they're random, it's a good idea to set a "seed", so you get the same layout more than once. 

- You have different layers (geoms) for edges vs nodes. 

  You don't have to include both---try commenting one or the other out.

```{r}
set.seed(123)

g_sw %>%
  ggraph(layout = "nicely") + 
  geom_edge_link(colour = "gray", alpha = .5) + 
  geom_node_point(size = 6, colour = "steelblue", alpha = .5) + 
  theme_graph() 

```

Let's add another node geom, to label the nodes with their names: 

```{r}
set.seed(123)
g_sw %>%
  ggraph(layout = "nicely") + 
  geom_edge_link(colour = "gray", alpha = .5) + 
  geom_node_point(size = 6, color = "steelblue", alpha = .5) + 
  geom_node_text(aes(label = name)) +
  theme_graph()
```

## Measuring communities

Finally, it's possible to algorithmically group nodes into *communities* in a variety of ways. Here's an example: 

```{r}
set.seed(123)

g_sw %>%
  mutate(community = as.factor(group_infomap())) %>%
  ggraph(layout = "nicely") + 
  geom_edge_link(colour = "gray", alpha = .5) + 
  geom_node_point(aes(color = community), size = 6, alpha = .5) + 
  geom_node_text(aes(label = name)) +
  theme_graph()

```

**1.5: Interpret this plot. If you've seen Star Wars, what are the two main groups here? (If you haven't, ask a friend.)**

# 2. US Congress Members on Twitter

Our second network is US Congress members on Twitter, from 2016: https://cdn.rawgit.com/pablobarbera/data-science-workshop/master/sna/03_challenge_1_solutions.html

```{r}
nodes <- read_csv("data/congress-twitter-network-nodes.csv", 
                  col_types = list(id_str = col_character()))

edges <- read_csv("data/congress-twitter-network-edges.csv", 
                  col_types = list(col_character(), col_character()))

g <- 
  graph_from_data_frame(d = edges, vertices = nodes, directed = TRUE) %>%
  as_tbl_graph()
```


**2.1: How many representatives and how many senators are in the data?**


**2.2: This is a *directed* graph. Why?**

## Senators by party affiliation

Let's plot just the senators: 

```{r}
set.seed(123)
g %>% 
  filter(chamber == "sen") %>%
  ggraph(layout = "nicely") + 
  geom_edge_link(colour = "gray", alpha = .1, width = .1) +
  geom_node_point(aes(color = fct_rev(party)), size = 2, alpha = .5) + 
  theme_graph() +
  labs(color = "Party affiliation")

```

## Labeling senators

Let's clean up the graph a bit. One senator isn't connected at all, so let's remove that one. Let's also label the senators from Washington.

```{r}
set.seed(123)
g %>%
  filter(chamber == "sen") %>%
  filter(!node_is_isolated()) %>%
  mutate(WA = (name %in% c("Patty Murray", "Maria Cantwell"))) %>%
  ggraph(layout = "nicely") + 
  geom_edge_link(colour = "gray", alpha = .1, width = .1) + 
  geom_node_point(aes(color = fct_rev(party)), size = 4, alpha = .4) + 
  geom_node_text(aes(label = name, filter = WA)) +
  theme_graph() +
  labs(color = "Party affiliation")
```

**2.3: Do you see evidence of polarization, i.e. of homophily by political party?**

If we convert the graph to an undirected graph, we can use a community detection algorithm on it: 

```{r}
set.seed(123)
g %>%
  filter(chamber == "sen") %>%
  filter(!node_is_isolated()) %>%
  to_undirected() %>%
  mutate(community = as.factor(group_louvain())) %>%
  ggraph(layout = "nicely") + 
  geom_edge_link(colour = "gray", alpha = .1, width = .1) + 
  geom_node_point(aes(color = community), size = 2,
                  alpha = .4) +
  theme_graph()
```

## Questions: House of Representatives

Do you see evidence of polarization in the House of Representatives too? Make a plot of representatives only. In order to make this work, *don't* plot the edges this time! If you do, your plot will be very slow. 

Also experiment with layouts other than "nicely". Force-directed layouts (like "kk" and "fr") can be especially useful. (In fact, "nicely" usually picks one of those for you.)

```{r}
set.seed(123)
# write your code here

```

## Questions: number of followers

You can use the `size` aesthetic to vary node size according to the number of followers. Do this for the Senate graph. 

Then, add labels to your plot for the top 5 senators by number of followers.

```{r}
set.seed(123)
# write your code here

```

# Challenge: Network measures

There are a variety of mathematical measures for characterizing properties of networks. `graph_assortativity` measures the amount of homophily in a graph. Calculate the assortativity by party for the House, the Senate, and Congress as a whole. 

Note: `graph_assortativity` needs to be used *within* the `mutate` function.

```{r}
# write your code here
```

