---
title: "Data transformation and visualization"
subtitle: "Soc 225: Data & Society"
author: "[PUT YOUR NAME HERE]"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

Today, we'll continue working with the Inside Airbnb data. This time, we'll use the full data set of listings for Seattle. 

Today's exercises are based off of Chapter 5 of *R for Data Science*: http://r4ds.had.co.nz/transform.html

# Setup

```{r}
library(tidyverse)
```

# Review: reading data

Once you've loaded the `tidyverse` package, the `read_csv` function will read data into R.

## Question 1: read the data

Read the `listings.csv` file from the `data/` folder into R and assign it to the variable name `airbnb_data` using the `<-` operator. 

```{r}
# write your code here
```

This data frame has more variables (= columns) and observations (= rows) than last week's data set. You can get an overview of the different columns using the `glimpse` function:

```{r}
glimpse(airbnb_data)
```

# Data transformation

## Pipes

First, let's practice piping! This is how we apply a series of transformations to data.

For example, to look at the first 6 unique values of the `neighbourhood` variable: 

```{r}
airbnb_data$neighbourhood %>% unique() %>% head()

# same as 
# head(unique(airbnb_data$neighbourhood))
```

## Verbs: filter

`filter` filters down to a subset of the data by **row**.

You can filter by values in one or more columns, using comparison operators like these:

- `==`: equal to
- `!=`: not equal to
- `>`, `<`, `>=`, `<=`
- `%in%`

This will give us the observations in the "Central Area" neighbourhood group:

```{r}
airbnb_data %>%
  filter(neighbourhood_group == "Central Area")
```

These are the 696 listings in the Central Area.

What about multiple criteria? For instance, if we only want to see listings in the Central Area with at least 10 reviews: 

```{r}
# `&` means "and", `|` means "or"
airbnb_data %>%
  filter(neighbourhood_group == "Central Area" & number_of_reviews >= 10)
```

There are 444 such listings.

### Question 2: How many listings?

How many listings are there in the UDistrict? Write code to filter the data set only to listings in the "University District" `neighbourhood_group`.

```{r}
# write your code here
```

Write the number of listings here:

## Verbs: select

`select` selects particular **columns**, by name: 

```{r}
airbnb_data %>%
  head(100) %>%
  select(id, name, neighbourhood_group, neighbourhood, 
         price, number_of_reviews) 
```

This is how I generated the data for Week 1.

You can use `-` to remove particular columns, too:

```{r}
airbnb_data %>%
  select(-latitude, -longitude)
```

## Verbs: arrange

`arrange` reorders a data set by one or more columns, in ascending or descending order.

Let's order the data by price, cheapest first:

```{r}
airbnb_data %>%
  select(id, name, host_name, 
         neighbourhood_group, neighbourhood, 
         price, number_of_reviews) %>%
  filter(price > 0) %>%
  arrange(price)
```

Most expensive first: 

```{r}
airbnb_data %>%
  select(id, name, host_name, 
         neighbourhood_group, neighbourhood, 
         price, number_of_reviews) %>%
  arrange(desc(price))
```

### Question 3: which listings have the most reviews?

Show the listings in the UDistrict that have the most reviews (`arrange` by `number_of_reviews` in `desc` order):

```{r}
# write your code here
```

# Data visualization

We'll start by plotting a single variable, using bar charts and histograms.

## Bar charts

There are three types of rooms: 

```{r}
unique(airbnb_data$room_type)
```

Let's plot how many there are of each using a bar chart. Bar charts are for *categorical* values. 

```{r}
airbnb_data %>% 
  ggplot(aes(x = room_type)) + 
  geom_bar()
```

## Histograms

What's the distribution of reviews that an Airbnb gets? We can see this using a histogram. Histograms are for *continuous* or numeric values.

```{r}
airbnb_data %>%
  ggplot(aes(x = number_of_reviews)) + 
  geom_histogram()
```

### Question 4: histogram of prices

Make a histogram of prices. When you do this, you'll see that it's hard to make sense of, because a few prices are really large.

To get a better look at the distribution of prices, filter out large values before plotting (e.g. `price <= 500`).


```{r}
# write your code here
```

# Challenge question: making a plot legible

The code below makes a bar plot of `neighbourhood_group`---but you can't read the neighborhood names! 

There are a couple ways to fix this. For instance, you could flip the x and y axes. Or, you could angle the text of the names. Try to figure out how to do one of those things, and modify the code below accordingly.

```{r}
airbnb_data %>%
  ggplot(aes(x = neighbourhood_group)) +
  geom_bar() 
```

# Just for fun: ggplot themes

The gray theme is the default, but ggplot has a variety of themes you can use. For example, I like the minimal theme better (`theme_minimal`). Type `?theme_gray` for a list. 

```{r}
airbnb_data %>%
  ggplot(aes(x = number_of_reviews)) + 
  geom_histogram() +
  theme_minimal() + 
  labs(title = "Distribution of reviews", 
       x = "number of reviews")

# for even more fun themes:
# install.packages("ggthemes")
# library(ggthemes)
# airbnb_data %>%
#   ggplot(aes(x = number_of_reviews)) + 
#   geom_histogram() +
#   theme_wsj()
```
