---
title: "Data transformation and visualization"
subtitle: "Soc 225: Data & Society"
author: "[PUT YOUR NAME HERE]"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

# Goals

- Begin exploring a data set on our own
- Master the pipe %>%
- Learn basic verbs from the tidyverse
- Produce basic visualizations

# 0: Lab Check-in

Today, we'll continue working with the Inside Airbnb data. This time, we'll use the full data set of listings for Seattle. 

Today's exercises are based off of Chapter 5 of *R for Data Science*: http://r4ds.had.co.nz/transform.html

**Question 0.1: Use the code block to produce the following code**

First, load the `tidyverse` package using the `library()` function.
Once you've loaded the `tidyverse` package, the `read_csv()` function will read data into R.
Read the `listings.csv` file from the `data/` folder into R and assign it to the variable name `airbnb_data` using the `<-` operator. 

```{r}

```

# 1: Exploring the Data

This data frame has more variables (= columns) and observations (= rows) than last lab's data set. You can get an overview of the different columns using the `glimpse` function:

```{r}
glimpse(airbnb_data) # display basic information about the data
```

Sometimes we'll just want to get the names of the columns. We can use the `names()` function to do that.


**Question 1.1: Display the column names of airbnb_data**

```{r}

```


**Question 1.2: Interpret the output from glimpse: what kind of information does it give you?**

# 2: Data transformation

## Pipes %>%

The pipe, `%>%`, is an extremely useful function that takes information from an element or function on the left and passes it to a function on the right. It lets you complete multiple operations in a single line. If you were making a whole bunch of peanut butter and pickle sandwiches, you could do it this way:

```
slices <- Cut(object = bread)
smeared <- Smear(object = slices, condiment = 'peanut_butter')
pickled <- Apply(object = smeared, what = 'pickles', method = 'carefully')
sandwich <- Close(object = pickled)
```

That's annoying because we're really doing all of those things to the same object. I don't really think of my partially made sandwich as 'smeared' or 'pickled'. Instead, I do all of those steps together, and keep working on the same object as I apply each transformation. That's what the pipe does for us in R. The same sandwich code would look like this:

```
sandwich <- 
  bread %>% 
  Cut() %>% 
  Smear(condiment = 'peanut_butter') %>% 
  Apply(what = 'pickles', method = 'carefully') %>% 
  Close()
```

You might notice that a key difference here is that the functions have lost their first arguments. That's becuase the pipe automatically takes the output from the left and uses it as the first arugment of the function to the right. So in the operation above, 'Cut' takes bread as its argument, 'Smear' takes Cut(bread) as its first argument, 'Apply' takes Smear(Cut(Bread), condiment = 'peanut_butter') as its first argument and so on.

Unfortunately, we don't have the technology yet to make real sandwiches this way. Instead, we'll be doing the same thing to data.

Let's practice piping!

For example, to look at the first 6 unique values of the `neighbourhood` variable: 

```{r}
# instead of
head(unique(airbnb_data$neighbourhood)) # show the top six unique neighborhoods

# we can write
airbnb_data$neighbourhood %>% unique() %>% head() # show the top six unique neighborhoods

```

**Question 2.1: Use the pipe to show the first 20 listing names**

```{r}

```


# 3: Verbs 

Now that we can pipe, we can start putting together different functions to perform multi-step analysis. In the tidyverse, we can think of most functions as 'verbs' whose names describe what they do. All* of them take a data frame as their first argument, so they're great for piping. We'll learn and practice using a few basic tidyverse verbs (mostly from the package 'dplyr' contined in the tidyverse), but there are many more. Remember that our course material offers a starting point! Coding is about facing new problems, making mistakes, using internet resources like Stack Exchange, and solving those problems (but your classmates and I are also here to help).

## filter

`filter` filters down to a subset of the data by **row**.

You can filter by values in one or more columns, using comparison operators like these:

- `==`: equal to
- `!=`: not equal to
- `>`, `<`, `>=`, `<=`
- `%in%`

Basically you can put anything into filter which will produce a logical vector (a list of TRUE and FALSE).

This will give us the observations in the "Central Area" neighbourhood group:

```{r}
airbnb_data %>% # take the airbnb data
  filter(neighbourhood_group == "Central Area") # filter to listings in the central area
```

These are the 696 listings in the Central Area.

What about multiple criteria? For instance, if we only want to see listings in the Central Area with at least 10 reviews: 

```{r}
# `&` means "and", `|` means "or"
airbnb_data %>% # take the airbnb data
  # filter to listings in the central area with more than ten reviews
  filter(neighbourhood_group == "Central Area" & number_of_reviews >= 10) 
```

There are 444 such listings.

**Question 3.1: How many listings?**

How many listings are there in the UDistrict? Write code to filter the data set only to listings in the "University District" `neighbourhood_group`.

```{r}

```

**Bonus: How many units are there in 'Other neighborhoods' that have a price less than $70 a night?**

```{r}

```


## select

`select` selects particular **columns** by name: 

```{r}
airbnb_data %>% # take airbnb data
  head(100) %>% # take the first 100 listings
  select(id, name, neighbourhood_group, neighbourhood, 
         price, number_of_reviews) # select only these columns
```

This is how we generated the data for Lab 1.

You can use `-` to remove particular columns, too:

```{r}
airbnb_data %>%
  select(-latitude, -longitude)
```

**Question 3.2: Use select to show only the prices**

```{r}

```

**Question 3.3: Use filter and select to display the prices and names of the listings with more than 10 reviews**

```{r}

```

## arrange

`arrange` reorders a data set by one or more columns, in ascending or descending order.

Let's order the data by price, cheapest first:

```{r}
airbnb_data %>% # take the airbnb data
  select(id, name, host_name, neighbourhood, 
         price, number_of_reviews) %>% # select only some columns
  filter(price > 0) %>% # exclude prices of $0 or less
  arrange(price) # arrange by ascending price order
```

Most expensive first: 

```{r}
airbnb_data %>% # take the airbnb data
  select(id, name, host_name, neighbourhood, 
         price, number_of_reviews) %>% # select only some columns
  filter(price > 0) %>% # exclude prices of $0 or less
  arrange(desc(price)) # arrange by descending price order
```

**Question 3.4: which listings have the most reviews?**

Show the listings in the UDistrict that have the most reviews:

```{r}

```

# 4: Data visualization

Now that we can adjust what data we're looking at with some verbs, we can start making visualizations. Not only do these look cool, but they're a great way to get a sense for big patterns in your data.

We'll start by plotting a single variable, using bar charts and histograms.

## Histograms

What's the distribution of reviews that an Airbnb gets? We can see this using a histogram. Histograms are for *continuous* or numeric values.

```{r}
airbnb_data %>%
  ggplot(aes(x = number_of_reviews)) + 
  geom_histogram()
```

**Question 4.1: Make a histogram of prices. When you do this, you'll see that it's hard to make sense of, because a few prices are really large.**

To get a better look at the distribution of prices, filter out large values before plotting (e.g. `price <= 500`).

## Bar charts

We can see that there are three types of rooms: 

```{r}
unique(airbnb_data$room_type)
```

Let's plot how many listings there are of from each neighborhood group using a bar chart. Bar charts are for *categorical* values. 

```{r}
airbnb_data %>% # take airbnb data
  ggplot(aes(x = neighbourhood_group)) + # start a plot and set the aesthetics with the x axis the neighborhood groups
  geom_bar() # make a bar chart
```

By adding another line of code, we can make the labels more legible:

```{r}
airbnb_data %>% # take airbnb data
  ggplot(aes(x = neighbourhood_group)) + # start a plot and set the aesthetics with the x axis the neighborhood groups
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + # adjust the labels so they're legible
  geom_bar() # make a bar chart
```

Finally, it's a good idea to give every plot a *title*, to tell people what it's about. 

```{r}
airbnb_data %>% 
  ggplot(aes(x = neighbourhood_group)) + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  geom_bar() +
  labs(title = "Airbnb listings in Seattle by neighborhood group")
```


** Question 4.2 Make a bar chart showing the number of listings for each room type **

```{r}
# consider adjusting the text angle as necessary

```

**Question 4.3 Would we want to use a bar chart or a histogram to visualize the number of reviews?**

**Question 4.4 Make the appropriate visualization**

```{r}

```


# Just for fun: ggplot themes

The gray theme is the default, but ggplot has a variety of themes you can use. For example, I like the minimal theme better (`theme_minimal`). Type `?theme_gray` for a list. 

```{r}
airbnb_data %>%
  ggplot(aes(x = number_of_reviews)) + 
  geom_histogram() +
  theme_minimal() + 
  labs(title = "Distribution of reviews", 
       x = "number of reviews")

# for even more fun themes:
# install.packages("ggthemes")
# library(ggthemes)
# airbnb_data %>%
#   ggplot(aes(x = number_of_reviews)) + 
#   geom_histogram() +
#   theme_wsj()
```

# Hints

1.2 Consider the this example output: $, 'id', '<int>', 2318, 4291, 5682, 6606
What do you think each element means?

2.1 You can use the same format as in the example, but you'll need to change 'neighbourhood' to 'name' and add 'n = 20' to the inside of the `head()` function.

3.3 First filter to listings with more than 10 reviews, then select the name and price

3.4 Try to `arrange` by `number_of_reviews` in `desc` order

3.5 Copy the code from the previous bar chart and change neighborhood_group to room_type.
