---
title: "Homework 3"
subtitle: "Soc 225: Data & Society"
author: "[PUT YOUR NAME HERE]"
date: "`r Sys.Date()`"
output: html_document
---

Write all code in the chunks provided!

Remember to unzip to a real directory before running everything!

To submit your assignment, knit to html and submit the output file.

Question one should be roughly analogous to what we've done in class. There are hints at the bottom of this document if you get stuck. If you still can't figure it out, go to google/stack exchange/ask a friend. Finally, email me or come to office hours :).

## Problem 1: Preparing and plotting data

### 1.1: Set up your environment

Set up your environment by: 

A. Loading the tidyverse  
B. Reading the Airbnb data: There's another new data set in the `data/` folder. This one has almost 10,000 cases and the census data by zipcode. These data are from New York City, not Seattle! 

```{r}

```
  
I've given you absolute populations and proportions for the racial composition of the zipcode for each listing. I've also made a variable called 'modal_race' which is the race with the largest proportion in that neighborhood. 

These variables are all in the last columns of the data set---you can try selecting them and using summary() to get a sense for what they contain.

### 1.2: Turn `price` into a number

`price` includes dollar signs, which means that R interprets it as a character. We want it to be a numeric variable instead. Turn `price` into a numeric variable in the chunk below.

There are a few ways to do this using `tidyverse` functions. See the hints below for some suggestions.

```{r}

```

### 1.3: Make a scatterplot

Use a scatter plot to compare how unit prices change with the proportion of a particular race. 

Bonus: try grouping by zipcode (in any fashion) for this plot

```{r}

```
  

### 1.4: Make a boxplot
  
Use the `modal_race` variable to plot a boxplot comparing race and price. You may have to look up how to make a boxplot in `ggplot2`---what geom do you need?

Bonus: try showing how this comparison differs by neighborhood group.

```{r}

```

### 1.5: Interpret your answer

Interpret your answer to 1.3. Check the hints if you need help.

*Your answer should be at least a few sentences here*

### Bonus: how did we make the data?

There's another file in the data folder, census.csv. Read it into R and have a look at it.

Download the full listings for New York City from Inside Airbnb, and see if you can join the Census data to it by zipcode using `left_join`. You'll have to filter out some wonky values for zipcode before you can merge. 

```{r}

```


## Problem 2: Plan your plots

You'll need to have **2 plots** to show the class for your final project. It helps to think about what you want to present and how you'll prepare and present that using your data.

In exploratory data analysis, you'll usually make more plots than you ultimately want to present. Because of that, we're asking you to make **3 plots** below, one more than necessary for the final project.

*For each of the 3 plots, provide:*
  
A. The purpose of the plot: what do you want people to understand when they see this?  

B. The type of plot: what geom functions will you use to present the plot? Why are those the best choices?  

C. Limitations/biases: What is missing from this presentation? Could someone get the wrong idea? What can you do to help limit the negative possibilities here?  

### Plot idea 1 

A.  
B.  
C.  

### Plot idea 2 

A.  
B.  
C.  

### Plot idea 3 

A.  
B.  
C.  
  
  
  
## Problem 3: Make the Plots 

Using your plans from Problem 2, produce the plots in R. Use as many code blocks as you need. Make sure to include proper titles and labels. 

Think about reducing chart junk and improving your data/ink ratio.
  
### Plot 1:  

```{r}

```

### Plot 2:  

```{r}

```

### Plot 3:  

```{r}

```

## Hints

1.2 

Use `mutate` for this. You can replace the original `price` variable, or name it something else. 

- `parse_number`, a function in the `readr` package, does a good job of converting currency to numbers on its own.

- `str_extract` with `pattern = "\\d+"`, then `as.numeric`, will extract numbers from a string, then convert the new (sub)string to a number.

- `str_remove_all`, with `pattern = "[\\$|,]"`, then `as.numeric`, will remove all dollar signs and commas.

1.5

Check out these resources if you're not sure about interpreting box plots:

https://magoosh.com/statistics/reading-interpreting-box-plots/

https://www.youtube.com/watch?v=oBREri10ZHk 

